### 2022-09-09 AISTATS2023/paper.tex
continued method description

@@

The terms \textit{uncertainty of explanation} and \textit{level of
  heterogeneous effects} refer to the same effect

@@

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
function \(f(\cdot)\) is a deterministic predictor, the local effect
is also a deterministic quantity. Global methods summarize the local
explanations into a single quantity. Therefore, ALE summarizes all the
local explanations at \(x_s\) by averaging the local effects across
all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
words, ALE globalizes the local explanations at \(x_s\), computing the
expected effect at \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the uncertainty of the local
effects at \(x_s\) as the variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the black-box function. It is important to clarify
that the variance is only a way to model the uncertainty. Other
statistical properties can also be used. Elaborate.

ALE computes the final effect at \(x_s\) by accumulating/integrating
the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE_accumulated_mean}). The
choice for \(x_{min}\) is not important, as it only affects the
centering of ALE plot along the vertical axis. Equivalently, we define
the accumulated uncertainty (variance) until the point \(x_s\), as the
integral of the variance of local effects:

\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent
In ALE plots the accumulated effect as absolute value is not an
interpretable quantity and is only used for better visual
interpretation\citep{Gromping2020MAEP}. The meaningful interpretation
is the effect at a specific point, i.e., what happens to the output
given a small change in the feature of interest. The integration takes
place only for making the visual interpretation easier, for example,
to spot larger intervals where the effect of a \textbf{local} change
is continuously constant. The same stands for the accumulated
uncertainty. The uncertainty has a meaninful interpretation as the
level of heterogeneous effects, only at a specific point. In the
experimental Section, we propose appropriate visualizations to
communicate to this effect to the user. However, the accumulated
uncertainty, i.e., the aggregated variance of the local effects, helps
us define a sensible objective for the interval spliting step. (TODO:
check here if the idea about whether the accumulated uncertainty is a
good metric for modeling the interaction strength)

@@

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
the limited instances of the training set. Unfortunately, it is
impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at
the granularity of a point \(x_s\), because The possibility to observe
a sample in the interval \([x_s - h, x_s + h]\) is zero, when
\(h \to 0\). Therefore, we are obliged split the axis \(x_s\) into a
sequence of larger non-overlaping intervals and estimate mean and
variance at the interval resolution. The mean effect at an interval
\([z_1, z_2)\) is the mean of the expected effects:

@@

\noindent
Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
is:

@@

As stated before, Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be
estimated from the set \(\mathcal{S}\) of the instances of the
training set with the \(s\)-th feature lying inside the interval,
i.e., \( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \}
\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be
approximated by:

@@

and the accumulated variance at the interval, Eq.~(\ref{eq:var_bin})
can be approximated by

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
\end{equation}

@@

ALE authors\citep{apley2020visualizing} proposed an approximation of
(\ref{eq:ALE_accumulated_mean}) by partitioning the axis of the
feature of interest into a sequence of non-overlaping bin and
computing a single effect in each bin. Later, (TODO: add citation to
us), proposed the DALE (Differential ALE) approximation that decouples
the two steps. The local effects are computed once using the partial
derivatives on the available instances of the training set,
i.e. \(\frac{\partial f}{\partial x_s}\), and the bin-spliting step,
simply allocates the precomputed local effects into bins, for
estimating the expected effect in each bin. The formula that estimates
the accumulate effects is:

\begin{equation}
  f^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x}
  \frac{1}{|\mathcal{S}_k|} \sum_{i:\xb^i \in \mathcal{S}_k}
  [f_s(\xb^i)] = \Delta x \sum_{k=1}^{k_x} \hat{\mu}_k
 \label{eq:DALE}
\end{equation}

>>>>

TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.

@@

@@

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are weigthed by the conditional distribution
\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the uncertainty of the local
effects at \(\mathcal{H}(x_s)\) as the variance of the local
explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In Section
(TODO), we propose appropriate visualizations for easier
interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the feature effect
at \(x_s\) is the accumulation of the averaged local effects from
\(x_{min}\) until \(x_s\), as show in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until the point \(x_s\), as the
integral of the variances of local effects:

\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent

The accumulated uncertainty is not a directly interpretable
quantity. It only helps us define a sensible objective for the
interval spliting step, as we discuss in Section TODO: adde ref.

@@

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point \(x_s\) is impossible, because the probability of observing a
sample inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). Therefore, we are obliged to split the axis of \(x_s\)
into a sequence of non-overlaping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:

@@

\noindent
Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
is defined as:

@@

Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
from the set \(\mathcal{S}\) of the dataset instances with the
\(s\)-th feature lying inside the interval, i.e.,
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:

@@

and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
approximated by

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
\end{equation}

@@

ALE, Eq.(\ref{eq:ALE_accumulated_mean}), is estimated by spliting the
axis \(x_s\) into a sequence of non-overlaping bins (TODO: add some
discussion for ALE, DALE):

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
\end{equation}

and

\begin{equation}
  \label{eq:ALE_accumulated_var_est}
  \hat{f}^{\mathtt{ALE}}_{\sigma^2}(x_s) = \sum_{k=1}^{k_x} \hat{\sigma}^2(z_{k-1}, z_k)
\end{equation}
